Parallelization of Push-based System for Molecular Simulation Data Analysis with GPU

### 2016-01-24 paper/paperSource.tex
paper changes...

Scientist gives the properties to simulation software (for example, Gromacs), runs the simulations, and finally get the output file. The output file must then be analyzed to produce certain results which may help him come up with certain consensus on original theoretical model that resulted in the molecular simulation system\cite{Frenkel:api01}. Gromacs is simulation software tool that helps scientist to run the actual simulation. It is a molecular dynamics package primarily designed for biomolecular systems such as proteins and lipids. \cite{Gromacs-online}. Besides the fact that it helps to generate the output files for the simulations, apparently it also helps to analyze the data itself, but the original problem is that it is not as optimized as it can be in order to analyze the data. Gromacs follows approach of pull-pased design, which means that for any given query (e.g. total mass or total charge, which are very similar type of 1-body queries without sophisticated selection) it will pull data separately and generate addition overhead wasting disk read I/O in order to come up with the result just for a single query. As it has been proposed in the previous paper, in order to remedy such issues, the push-based design does exactly the opposite, where instead of loading the data on demand for each query, the queries are batched into a network, then the entire dataset is loaded chunk by chunk pushing it through the network which has its internal relationships and dependencies amonth the queries. In this case, since scientific simulation data is run once with specific physical properties, it is never modified, thus, on continuation, it will only append, which means that the processing can also be run on appended frames. \cite{mainPaper}
This type of approach has already been revised by other systems \cite {DataPath,Volcano,Qpipe} as of reusing loaded data through queries produced \cite {Candea,PredictablePerformance,CooperativeScans}.

@@

\subsection{Problem Statement}
As mentioned earlier, in order to describe the scientific system,
scientist must look into/analyze the critical statistical properties
of the data produced by the simulation. These properties are usually
computed through series of queries that are being executed on the
whole, or some selection of the MS data. Most of the systems that
are widely used today across many scientific fields for analysis of
the MS data are part of a software systems that also run the
simulation. Such examples include GROMACS~\cite{GROMACS4},
VMD~\cite{VMD}, MDAnalysis~\cite{MDAnalysis}, Wordom~\cite{wordom},
MD-TRACKS~\cite{MDtracks}, SimulaidOne~\cite{Simulaid},
Charmm~\cite{CHARMM}. Once the simulation is run, the system
produces a flat file with all the measurements of each particle.
These systems take each query as a user's input and apply them to
the data uploaded from these flat files (called trajectories in
GROMACS). The biggest problem/performance issue that these system
impose is the fact that for every new query issued by the user, the
system has to load a significant part of the dataset into main
memory before executing the query. Such pull-based design involves
random I/Os that considerably affecting the data throughput in the
system. And while the use of an index-based scan might seem better
option to a sequential scan, this is the case only when the query
accesses small portion of the data. All of this gives rise of
redundant and high I/O traffic in the process of scientific data
analysis in pull-based type systems. This adds greatly to the
overall performance time of the analysis of the MS system. Having in
mind the volume of data produced by the MS in a single frame (in
gigabytes), and we add the number of frames simulated (tens of
thousands) the total volume of data to be analyzed is of huge
magnitude. So to have to load big chunks of this data from disk to
the main memory every time a query is being executed degrades the
overall system's performance greatly. In this paper we design and
implement a push-based type system which allows high-throughput data
analysis in the process of scientific discovery.

\subsection{Our Approach}
Considering the fact that the running time of the most queries used
in the scientific world to analyze the MS is minuscule in comparison
to the loading time of the MS data, it is safe to say that the
scientific world would benefit from a system that can remedy the
aforementioned problem. Our idea is to build a system that would
load the MS data from hard disk in the main memory only once and
then execute as many queries as needed/wanted on that data, without
the need of reloading the same data over and over again. Considering
the volume of data in a single MS frame as well as the number of
frames produced during a single simulation, this type of system can
save a lot of time for MS analysis. The system would contain many,
most often used queries that can also be run as user's input. These
queries would be pre-programmed as separate modules in our system,
but they would be able to take certain attributes, like data
selection for instance, as user's input. Once the system is run, all
(or a selection) of these queries would be executed against the MS
data or certain selection. The system would act as a type of
push-based system, essentially pushing the loaded data to all the
queries. Once we have the data in the memory, the reading can loop
as many times as the queries need it. This is useful for some
non-streaming functions/queries that cannot be computed in a single
run of the data. Such one query is the SDH. But we have also come up
with a way to resolve this in a way, by creating a quad-tree based
database for this query. This quad-tree (we called Density Map (DM))
is being populated while the data is being loaded into main memory.
After the DM is populated, we can compute the SDH with much faster
algorithm than the brute-force method used everywhere else.

>>>>

Scientist gives the properties to simulation software (for example, Gromacs), runs the simulations, and finally get the output file. The output file must then be analyzed to produce certain results which may help him come up with certain consensus on original theoretical model that resulted in the molecular simulation system\cite{Frenkel:api01}. Gromacs is simulation software tool that helps scientist to run the actual simulation. It is a molecular dynamics package primarily designed for biomolecular systems such as proteins and lipids. \cite{Gromacs-online}. Besides the fact that it helps to generate the output files for the simulations, apparently it also helps to analyze the data itself, but the original problem is that it is not as optimized as it can be in order to analyze the data. Gromacs follows approach of pull-based design, which means that for any given query (e.g. total mass or total charge, which are very similar type of 1-body queries without sophisticated selection) it will pull data separately and generate addition overhead wasting disk read I/O in order to come up with the result just for a single query. As it has been proposed in the previous paper, in order to remedy such issues, the push-based design does exactly the opposite, where instead of loading the data on demand for each query, the queries are batched into a network, then the entire dataset is loaded chunk by chunk pushing it through the network which has its internal relationships and dependencies amonth the queries. In this case, since scientific simulation data is run once with specific physical properties, it is never modified, thus, on continuation, it will only append, which means that the processing can also be run on appended frames. \cite{mainPaper}
This type of approach has already been revised by other systems \cite {DataPath,Volcano,Qpipe} as of reusing loaded data through queries produced \cite {Candea,PredictablePerformance,CooperativeScans}.

@@

\subsection{Problem Statement}
Simulation software systems, in general, follow the same methodology of running and storing simulation data. The simulation software system examples are: Gromacs \cite{GROMACS4}, VMD\cite{VMD}, MDAnalysis~\cite{MDAnalysis}, Wordom~\cite{wordom}, MD-TRACKS~\cite{MDtracks}, SimulaidOne~\cite{Simulaid}, Charmm~\cite{CHARMM}. In the type of simulations brought up as examples above, the flow of the data is the following. Once the simulation is run, the output files are contained as trajectory files with descriptors (they contain information about space dimensions, number of atoms and frames, etc.) that can be easily transposed into simple flat files containing the physical properties atom by atom, frame by frame, which are consequently read and processed by the proposed push-based system. Since we have certain amount of queries needed to be run on given simulation data, generating high I/O traffic followed by design of pull-based system is not considerable. The approach proposed in the previous paper is very good in terms of performance in comparison with the original pull-based system\cite{mainPaper}. The problem is still that some of the queries processed by those means are still improvable, especially taking into consideration the fact that in the used previous works for calculation of 2-body functions, which take the biggest time for processing~\cite{ytu:icde09, EDBT12}, we might have some error bounds, which might be unacceptable for certain simulation analysis where sacrifice on loss of data is unbearable. Besides it, although we might have huge performance boost on specific data structure as density map, we still have certain expectation from data nature (such as its uniformity), thus, it makes sense to have desire to simplify the processing and still having pure computation based on any values, considering that the queries should be available to be pre-programmed by user in a separate module of the code.

\begin{figure}
 \centerline{ \includegraphics[width=1\columnwidth]{images/speedup838K-pomalo-za-4-eps-converted-to.pdf} }
 \caption{ Speed up over different levels of atom selection for 838,400 atoms \cite{mainArticle} }
 \label{fg:sample_estimation_old_paper}
\end{figure}

\subsection{Our approach with improvement}

Reading the data from files is a huge overhead, thus, speedup of push-based system in general is pretty significant over pull-based system in our given case. 
It's been demonstrated in the previous paper with different amount of atoms, frames, and workloads \cite{mainPaper}. Since we have to load the data every time there is a different query versus push-based system loads a chunk of data once, it is very noticable how the framework \cite{mainPaper} contributes to efficiency of analysis of simulation data. You can observe some estimation samples in Figure \ref{fg:sample_estimation_old_paper}

On the other hand, some of the queries may be very slow due to their nature on a sequential type of computation,  especially, 2-body functions. Since the nature of the data that comes with simulation is basically physical properties of atoms, thus, there is a lot of computation that involves independent primitive mathematical operations, which makes it a perfect problem for parallelism.

Considering the fact that the running time of the most queries used
in the scientific world to analyze the MS is minuscule in comparison
to the loading time of the MS data, it is safe to say that the
scientific world would benefit from a system that can remedy the
aforementioned problem. Our idea is to build a system that would
load the MS data from hard disk in the main memory only once and
then execute as many queries as needed/wanted on that data, without
the need of reloading the same data over and over again. Considering
the volume of data in a single MS frame as well as the number of
frames produced during a single simulation, this type of system can
save a lot of time for MS analysis. The system would contain many,
most often used queries that can also be run as user's input. These
queries would be pre-programmed as separate modules in our system,
but they would be able to take certain attributes, like data
selection for instance, as user's input. Once the system is run, all
(or a selection) of these queries would be executed against the MS
data or certain selection. The system would act as a type of
push-based system, essentially pushing the loaded data to all the
queries. Once we have the data in the memory, the reading can loop
as many times as the queries need it. This is useful for some
non-streaming functions/queries that cannot be computed in a single
run of the data. Such one query is the SDH. But we have also come up
with a way to resolve this in a way, by creating a quad-tree based
database for this query. This quad-tree (we called Density Map (DM))
is being populated while the data is being loaded into main memory.
After the DM is populated, we can compute the SDH with much faster
algorithm than the brute-force method used everywhere else.
